---
title: "Line_graph"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
# Allows us to read-in csv files
library(readr) 
# For data manipulation
library(dplyr) 
# For regular expression operations 
library(stringr) 
# Used tp create interactive visualisations
library(plotly)
```

# Dataset 3

This dataset includes sexual identity estimates by gender from 2010 to 2014. This is presented at a UK level, and broken down by England, Wales, Scotland and Northern Ireland. I wanted this guide to include a demo of how to make interactive line graphs with gender identity data, but unfortunately given this is only the first year that the ONS has collected this data that was not possible. So I found a dataset from 2015 which involves experimental statistics that have been used in the Integrated Household Survey. For more info, you can check out this [ONS link](https://www.ons.gov.uk/peoplepopulationandcommunity/culturalidentity/sexuality/datasets/sexualidentitybyagegroupbycountry). 

```{r}
# Load in dataset

df3 <- read_csv('../Data/cleaned_sexuality_df.csv')
```

```{r}
# Brief glimpse at underlying data structure

head(df3, 10)
```

## Data cleaning

When I first found this dataset it was very messy and formatted terribly, so I performed some cleaning on it in a separate jupyter notebook, to save cluttering this one and distracting from the main tutorial. If you'd like to see how I cleaned it up, please see the ['Data_cleaning_sexuality.ipynb'](Data_cleaning_sexuality.ipynb) notebook. 

## Data pre-processing

The only pre-processing we're going to do is subset our data by country, and also create 2 separate datasets for Gender = Men and Gender = Women. I'll explain why this step is needed soon. 

```{r}
# Filter dataset to focus on England
england_df <- df3 %>%
  filter(Country == 'England')
```

```{r}
# Let's check it worked.. 

unique(england_df$Country)
```

```{r}
# Further filter data for each gender

men <- england_df %>% filter(Gender == "Men")
women <- england_df %>% filter(Gender == "Women")

# Let's check it worked

unique(men$Gender)
unique(women$Gender)
```


## Interactive linegraph

Creating a simple line graph in plotly is pretty easy, but where plotly struggles (in R) is in handling facet plots. A facet plot is a type of visualisation that divides data into subplots based on categorical variables. What I'd like to do is create a facet plot of sexuality percentages in England (2010-2014) with individual subplots for our two genders. This is achieved easily in Python due to the plotly.express module, which provides a simple way to create facet plots. Unfortunately, we'll have to go through a bit more of a longwinded route, where we'll manually create our individual plots for each gender, then combine them using the subplot function. Also, plotly.express automatically manages legends to ensure they're unified across facets, but R's plotly requires that we manually sync up these legends. Womp womp. Let's get to it. 



```{r}
# Create individual plot for each gender

# Create plots for each gender
men_plot <- plot_ly(men, 
                    x = ~Year, 
                    y = ~Percentage, 
                    color = ~Sexuality, 
                    type = 'scatter', 
                    # mode used to make sure our data points are connected by lines across the years
                    mode = 'lines+markers', 
                    hoverinfo = 'text',
                    text = ~paste("Year:", Year, "<br>Percentage:", Percentage, "<br>Sexuality:", Sexuality),
                    # legendgroup parameter ensures that data points relating to the same category are synced across plots
                    legendgroup = ~Sexuality,
                    # showlegend parameter set to TRUE only for this plot to avoid duplicate legends
                    showlegend = TRUE) %>%
  layout(xaxis = list(title = 'Year', tickvals = 2010:2014, ticktext = 2010:2014),
         yaxis = list(title = 'Percentage'),
         # Here we add an annotation to the graph to label the first subplot "Men"
         # Setting xref and yref to 'paper' simply means the annotation won't move if we zoom in or out
         annotations = list(
           list(x = 0.5, y = 1.05, text = "Men", showarrow = FALSE, xref='paper', yref='paper')))


women_plot <- plot_ly(women, 
                      x = ~Year, 
                      y = ~Percentage, 
                      color = ~Sexuality, 
                      type = 'scatter', 
                      mode = 'lines+markers', 
                      hoverinfo = 'text',
                      text = ~paste("Year:", Year, "<br>Percentage:", Percentage, "<br>Sexuality:", Sexuality),
                      legendgroup = ~Sexuality,
                      showlegend = FALSE) %>%
  layout(xaxis = list(title = 'Year', tickvals = 2010:2014, ticktext = 2010:2014),
         yaxis = list(title = 'Percentage'),
         annotations = list(
           list(x = 0.5, y = 1.05, text = "Women", showarrow = FALSE, xref='paper', yref='paper')))

# Let's take a look at one of these graphs

women_plot
```

```{r}
# Combine individual plots using subplot
# Within subplot, define number of rows, make sure share same x axes and both axes titles
fig5 <- subplot(men_plot, women_plot, nrows = 2, shareX = TRUE, titleX = TRUE, titleY = TRUE) %>%
  layout(
    title = list(
      text = 'Sexuality Percentages by Gender in England (2010-2014)', 
      y = 0.98,  # Move the title higher up
      x = 0.5,   # Center the title
      xanchor = "center",
      yanchor = "top"
    ),
    margin = list(t = 100),  # Add space at the top for the title
    height = 800,
    width = 1000
  )

fig5
```
